A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-08 text/manuscript.Rmd
add fig s4, table st1, incomplete fig 4 - still needs work

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(umap)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
```

@@

@@

selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_mut, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d"), meausure.vars = tumor_suppressor$external_gene_name) %>%
  filter(value != 0)

@@

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]
![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(umap)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
```

@@

```{r figure-4-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_2d_3d_coors.tsv") ## load in data
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene_data.Rda")

## Parameters ##
filtered_class <- c(6,7) ## filtering for only class 6 and 7 samples
n_path <- 10 ## looking at the top10 path

## Data Generation ##
class_spec <- tumor_clusters %>% dplyr::select(V1, knn_clust) %>% filter(knn_clust %in% filtered_class) %>% .[,.$V1] ## sample_id for class specific

toppath <- gdc_reactome_path_analysis %>% .[, names(.) %in% class_spec] %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top10 mutated pathways

genes_in_path <- gene2level2_lookup %>% .[, match(pathname2level2$PATH_ID, names(.))] %>% ## match path ids
  'colnames<-' (pathname2level2$PATH_NAME) %>% ## replace with path_names
  dplyr::select(c(toppath)) %>% ## filter by top 10 pathways
  rownames_to_column(var = "ensembl_gene_id") %>% left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl gene id with gene name
  .[rowSums(.[, -1])>0,]

df <- upset(genes_in_path, 
      sets = names(genes_in_path),
      mb.ratio = c(0.55, 0.45), 
      order.by = "freq", 
      point.size = 3.5,
      line.size = 2.0,
      mainbar.y.label = "Gene Intersection",
      sets.x.label = "Total Gene Per Pathway",
      text.scale = c(2.5, 2.4, 2, 2.1, 2.4, 1.5),
      number.angles = 0)

make_combinations <- function(x) {
  l <- length(x)
  mylist <- lapply(1:l, function(y) {
    combn(x, y, simplify = FALSE)
  })
  mylist
}

path_combo <- make_combinations(names(genes_in_path)) ## possible pathway combinations from 1 to 10

top10path <- genes_in_path %>% rowSums() %>% as.data.frame() %>% ## getting the sums of genes in pathway
  'colnames<-' ("sum_top10") %>% ## rename col
  rownames_to_column(var = "gene_name") ## make rownames a col

path_match_in_set <- list() ## making a list for matchings in sets

for(n in 1:10) {
  message(n)
  path_combo_set <- path_combo[[n]] ## selecting combination set (1 combo, 2 combo, 3 combo, etc)
  path_match_within_set <- list() ## making a list for matching within sets
  for(p in 1:length(path_combo_set)) {
    message(p)
    path_name <- path_combo_set[[p]] ## selecting specific combination pathway names
    spec_combo <- genes_in_path %>% dplyr::select(all_of(path_name)) ## select pathway combo
    if(length(spec_combo) > 1) {
      spec_combo <- spec_combo %>% rowSums() %>% as.data.frame() %>% ## rowSums the pathway to get the gene total membership
        'colnames<-' ("combo") %>% filter(combo == n) ## rename col and filter for combo that is n (n combo)
    }
    if(length(spec_combo) == 1) {
      names(spec_combo) <- "combo"
    }
    spec_combo <- spec_combo %>% rownames_to_column(var = "gene_name") %>% left_join(., top10path, by = "gene_name")
    spec_combo$match <- spec_combo$combo == spec_combo$sum_top10
    any_matches <- as.character(unique(spec_combo$match)) ## are there any matches?
    if(!"TRUE" %in% any_matches) { ## if no matches then skip
      next
      }
    if("TRUE" %in% any_matches) { ## if are matches then do this
      path_match_within_set[[p]] <- spec_combo %>% filter(match == "TRUE") %>% dplyr::select(gene_name) %>% ## filter for trues and select only gene names
        mutate(path_combo = paste0(paste(path_name, collapse=", "))) ## insert combo type
    }
  }    
  if(length(path_match_within_set) > 0) {
    path_match_in_set[[n]] <- path_match_within_set %>% do.call(rbind, .) %>% as.data.frame()
  }
}
path_match_in_set <- path_match_in_set %>% do.call(rbind, .) %>% as.data.frame() ## combine the list together

save(path_match_in_set, file = "path_match_in_set.Rda")

### getting clustering order from pheatmap
most_genes <- path_match_in_set %>% .$path_combo %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% ## arranging pathway combination based on most genes involved
  'colnames<-' (c("path_combo", "Freq")) %>% dplyr::select(path_combo) %>%
  rownames_to_column(var = "order")

gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
  arrange(order) %>% .$gene_name ## getting the gene name order

## now need to get the gene mutation data and tumor samples 
load("../data/gdc_tumor_coding_complete.Rdata")

gdc_mut_count_patient <- missense + splice + frameshdel + frameshins + utr3 + utr5 + nonstop + nonsense ## add up all the mutation count
rm(missense, splice, frameshdel, frameshins, utr3, utr5, nonstop, nonsense, gdc_tumor_coding_complete)
gc()
gdc_mut_count_patient[gdc_mut_count_patient > 0] <- 1 ## booleanize data, anything greater than 1 is equal to 1

filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
  column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
  .[rownames(.) %in% gene_order, names(.) %in% class_spec] ## filter by genes in top10 path of filtered class and tumor samples associated with it
rm(gdc_mut_count_patient)
gc()

save(filtered_mut_data, gene_order, file = "filtered_mut_data.Rda")
pheatmap(filtered_mut_data[gene_order,],
         show_rownames = T,
         show_colnames = F,
         treeheight_row = 0,
         treeheight_col = 0,
         cellheight = 5,
         cellwidth = 1,
         cluster_cols = T,
         cluster_rows = F,
         #gaps_row = file_dims,
         #cutree_cols = 6,
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2",
         fontsize = 10,
         legend = T,
         filename = "fig4.png")

## stopped here ## ignore bottom using as reference
path_match_in_set
cancer_heatmap <- pheatmap(cancer_path[pathway_dendro,], ## using pathway row order 
                          show_rownames = F,
                          show_colnames = F,
                          cellheight = 1,
                          cellwidth = 0.5,
                          cluster_cols = T,
                          cluster_rows = F,
                          annotation_legend = F,
                          color = c("grey", "firebrick3"),
                          clustering_method = "single", ## ordering is single linkage
                          legend = T)
  cancer_order <- cancer_heatmap$tree_col$order ## cancer order
  cancer_path <- names(cancer_path) ## getting the sample_id original order 
  cancer_dendro[[n]] <- cancer_path[cancer_order] ## reorder cancer by heatmap order
  rm(cancer_info, cancer_path)
  gc()
}
dev.off()

cancer_disrupt_heatmap <- left_join(cancer_disrupt_heatmap, cancer_order, by = "sample_id") ## add in cancer type information for faceting
save(cancer_disrupt_heatmap, file = "cancer_disrupt_heatmap.Rda") 
```

```{r figure-4}

@@

selected_mut <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_mut, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
  filter(value != 0)

@@

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]

<div style="text-align: right"> Supplementary Figure S4 Data </div>
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway

selected_path <- gdc_reactome_path_analysis %>% t() %>% as.data.frame() %>% rownames_to_column(var = "V1") ## prepare pathway for left_join

selected_path <- umap_cancertype %>% dplyr::select(V1, x_2d, y_2d) %>% ## select relevant columns
  left_join(., selected_path, by = "V1") %>%
  melt(data = ., id.vars = c("V1", "x_2d", "y_2d")) %>%
  filter(value != 0)

save(selected_path, file = "selected_path.Rda")
```

```{r supplementary-figure-s4, out.width = '90%'}
load("../data/selected_path.Rda") ## load data
load("../data/gdc_reactome_path_analysis.Rda")
umap_cancertype <- fread("../data/umap_2d_3d_coors.tsv")
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

n_path <- 377 ## number of pathways we want to see

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## order pathway by most to least mutated pathway.. needed for facet

ggplot(selected_path, aes(x = x_2d, y = y_2d, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = x_2d, y = y_2d), colour = "grey", size = 0.5) +
  geom_point(size = 0.5) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = toppath),
                      ncol = 5) ## facet by pathways

![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]

<div style="text-align: right"> Supplementary Figure ST1 Data </div>
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

## Parameters ##
n_path <- 377 ## how many top pathways do you want to see. 1 = top path, 2 = top path and second most
mutated_gene_position <- c(1,2,3) ## what gene do you want to see. 1 = top gene, 2 = second most, 3 = third
show_perc <- TRUE ## show perc for mutated gene

toppath <- gdc_reactome_path_analysis %>% rowSums() %>% as.data.frame() %>% rownames_to_column(var = "path_name") %>% arrange(desc(.)) %>% .[1:n_path,] %>% .[, "path_name"] ## getting the top mutated pathways

id_to_names <- gene2level2_lookup %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id) %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "PATH_ID") %>% ## convert path_id to path_name
  left_join(., pathname2level2, by = "PATH_ID") %>%
  remove_rownames() %>% column_to_rownames(var = "PATH_NAME") %>%
  dplyr::select(-PATH_ID) %>%
  t() %>% as.data.frame()

gene_mutation <- gdc_gene_exp_adj %>% rownames_to_column(var = "ensembl_gene_id") %>% ## convert ensembl to gene name
  left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>%
  remove_rownames() %>% column_to_rownames(var = "external_gene_name") %>%
  dplyr::select(-ensembl_gene_id)

table_st1 <- list() ## list for forloop

for(n in toppath) {
  message(n)
  ## finding top mutated gene in pathway
  gene_in_path <- id_to_names %>% .[,which(names(.) == n)] %>% as.data.frame() %>% ## select pathway
    'rownames<-' (rownames(id_to_names)) %>% 
    filter(. == 1) %>% ## filter for genes that are in that pathway
    rownames(.)
  top_mut_gene <- gene_mutation %>% .[rownames(.) %in% gene_in_path, ] %>% ## only include genes in paths
      rowSums() %>% as.data.frame() %>% 'colnames<-' ("obs_count") %>% ## sum the pathway mutation
      mutate(n_sample = length(gene_mutation),
             perc_mut = round((obs_count / length(gene_mutation)) * 100, digits = 1)) %>% 
      rownames_to_column(var = "gene") %>% arrange(desc(obs_count))
  top_gene_in_path <- list()
  for(p in mutated_gene_position) {
    message(p)
    if(show_perc == TRUE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p], " (", top_mut_gene$perc_mut[p],"%)")  
    }
    if(show_perc == FALSE) {
    top_gene_in_path[[p]] <- paste0(top_mut_gene$gene[p])  
    }
  }
  top_gene_in_path <- top_gene_in_path %>% do.call(rbind, .) %>% as.data.frame() %>% ## rbind list
   'rownames<-' (mutated_gene_position) %>% t() %>% as.data.frame()
  ## calculating the percent mutated samples
  spec_path <- data.frame(Pathway = n,
                          Percent_Mutated = round(gdc_reactome_path_analysis %>% .[n,] %>% sum() / length(gdc_reactome_path_analysis) * 100, digits = 1))
  table_st1[[n]] <- cbind(spec_path, top_gene_in_path)
}
table_st1 <- as.data.frame(do.call(rbind, table_st1))
save(table_st1, file = "table_st1.Rda")  
rm(list = ls(all.names = T))  
gc()
```

```{r supplemental-table-st1}
load("../data/table_st1.Rda")

table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T, color = "black")
```

![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]

![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)
